Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-26 inference.tex
edits

constructively by a meta function $\stabilize{\cdot}$, which accepts

@@

Due to the presence of bind ($\bind$) in $\SL$, a set expression $s$
may contain free variables introduced by an enclosing binder. For
instance, consider the $\SL$ expression $s_1\bind(\lambda x. \{ y\,|\,
y=x+1\})$, where $s_1$ is an integer set (expression). The
subexpression $\{ y\,|\, y=x+1\}$ (call it $s_2$) contains $x$ as a
free variable. In such cases, the predicate associated with the
subexpression should also be indexed by its free variables, so that a
unique set exists for each instantiation of the free variables. Thus,
the predicate ($\G$) associated with the subexpression from the
above example should be $\lambda\v_1.\lambda\v_2.~\v_2 = \v_1 + 1$, so
that the set $\G\; x_1$ is different from the set $\G\; x_2$ for
distinct $x_1,x_2 \in s_1$. Intuitively, the bind expression
$s_1\bind(\lambda x. \{ y\,|\, y=x+1\})$ denotes the set
$\bigcup\limits_{x\in s_1}\G\;x$.

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
(possible) free variables of $s$, and thus its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation ``$\in$'' to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
(denoted $\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
A set union expression is encoded similarly.

@@

Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
is extended only in the encoding rules for $\bind$ and $\existsl$
expression. Since $\existsl$ is a result of a particular choice of the
weakening semantics (Sec.~\ref{sec:inference}), let us ignore it to
focus on the core set expression language. This leaves $\bind$ as the
only case where $\nubar$ is extended.  Since an occurance of $\bind$
adds one to $|\nubar|$, if $s$ is a subexpression of a top-level
$\bind$ expression (like $\F_U(\stl,\stg)$ from the above example),
then the prenex signature of $\phi_s$ is $\forall^2\exists$.
Furthermore, if the subexpressions of $s$ are not $\bind$ expressions,
then none of the $\phi_i$'s are quantified, and the prenex signature
of $\phi = \bigwedge_i\phi_i \wedge \phi_s$ remains
$\forall^2\exists$.  The fragment of FOL that admits formulas with
prenex signatures of form $\forall^2\exists^*$ is called
Godel-Kalmar-Schutte ({\sf GKS}) fragment, which is known to be
decidable. The language of encoding, however, is a combination of {\sf
GKS} with (a). $\Prop^0$, the theory from which quantifier-free
propositions ($\varphi$) that encode object language expressions are
drawn, and (b).  $\Prop^1$, the theory from which invariants ($I$) are
drawn. Thus the encoding is decidable if the combination of ${\sf GKS}
+ \Prop^0 + \Prop^1$ is decidable.

To formally state the above property, let $\SL[n,\Prop^0,\Prop^1]$
denote the core set expression language that admits $n$ nested $\bind$
expressions, and allows program's (boolean) expressions ($e$), and
invariants ($\I$) to be drawn from theories $\Prop^0$ and $\Prop^1$,
respectively. Then, we have the following result:
\begin{theorem}
  $\SL[0,\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
The $n=0$ case denotes the lack of nested bind expressions. Since
nested \C{foreach} statements are the only program constructs in
$\txnimp$ that lead to nested bind expressions, the $n=0$ case
essentially denotes the encoding language for $\txnimp$ programs
without nested \C{foreach} loops. Observe that the \C{new\_order}
transaction (Fig.~\ref{fig:new_order_code}) satisfies this constraint.
Indeed, the restricted language without nested \C{foreach} loops is
nonetheless expressive enough to cover all TPC-C transactions.

>>>>

constructively by a meta-function $\stabilize{\cdot}$, which accepts

@@

Due to the presence of bind ($\bind$) in $\SL$, a set expression $s$
may contain free variables introduced by an enclosing binder. For
instance, consider the $\SL$ expression $s_1\bind(\lambda x. \{ y\,|\,
y=x+1\})$, where $s_1$ is an integer set (expression). The
subexpression $\{ y\,|\, y=x+1\}$ (call it $s_2$) contains $x$ as a
free variable. In such cases, the predicate associated with the
subexpression should also be indexed by its free variables so that a
unique set exists for each instantiation of the free variables. Thus,
the predicate ($\G$) associated with the subexpression from the
above example should be $\lambda\v_1.\lambda\v_2.~\v_2 = \v_1 + 1$, so
that the set $\G\; x_1$ is different from the set $\G\; x_2$ for
distinct $x_1,x_2 \in s_1$. Intuitively, the bind expression
$s_1\bind(\lambda x. \{ y\,|\, y=x+1\})$ denotes the set
$\bigcup\limits_{x\in s_1}\G\;x$.

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta-function $\G$ is a predicate indexed by the
(possible) free variables of $s$, and thus its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation ``$\in$'' to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
(denoted $\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
A set union expression is encoded similarly.

@@

Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
is extended only in the encoding rules for $\bind$ and $\existsl$
expressions. Since $\existsl$ is a result of a particular choice of the
weakening semantics (Sec.~\ref{sec:inference}), let us ignore it for now to
focus on the core set expression language. This leaves $\bind$ as the
only case where $\nubar$ is extended.  Since an occurrence of $\bind$
adds one to $|\nubar|$, if $s$ is a subexpression of a top-level
{\sf bind} expression (like $\F_U(\stl,\stg)$ from the above example),
then the prenex signature of $\phi_s$ is $\forall^2\exists$.
Furthermore, if the subexpressions of $s$ are not {\sf bind} expressions,
then none of the $\phi_i$'s are quantified, and the prenex signature
of $\phi = \bigwedge_i\phi_i \wedge \phi_s$ remains
$\forall^2\exists$.  The fragment of FOL that admits formulas with
prenex signatures of form $\forall^2\exists^*$ is called the
Godel-Kalmar-Schutte ({\sf GKS}) fragment~\cite{...}, which is known to be
decidable. The language of encoding, however, is a combination of {\sf
GKS} with (a). $\Prop^0$, the theory from which quantifier-free
propositions ($\varphi$) that encode object language expressions are
drawn, and (b).  $\Prop^1$, the theory from which invariants ($I$) are
drawn. Thus the encoding is decidable if the combination of ${\sf GKS}
+ \Prop^0 + \Prop^1$ is decidable.

To formally state the above property, let $\SL[n,\Prop^0,\Prop^1]$
denote the core set expression language that admits $n$ nested {\sf bind}
expressions, and allows program (Boolean) expressions ($e$), and
invariants ($\I$) to be drawn from theories $\Prop^0$ and $\Prop^1$,
respectively. Then:
\begin{theorem}
  $\SL[0,\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
The $n=0$ case denotes the lack of nested {\sf bind}
expressions. Since nested \C{foreach} statements are the only program
constructs in $\txnimp$ that lead to nested bind expressions, this
instantiation denotes the encoding language for $\txnimp$ programs
without nested \C{foreach} loops. Observe that the \C{new\_order}
transaction (Fig.~\ref{fig:new_order_code}) satisfies this constraint.
Indeed, this restricted language is nonetheless expressive enough to
cover all of the benchmarks we have considered.
